\vspace{3cm} \Huge{Optimizing SasView models} \\ \Large{Enhancing performance by rewriting kernels from OpenCL/OpenMP to Futhark}

### 2017-11-07 report/draft.tex
Pøls!

\newcommand{\sasmodels}{\textit{SasModels}}
\newcommand{\sasview}{\textit{SasView}}
\newcommand{\iq}{\textit{I(q)}}
\begin{document}

@@

\subsection{Defining a Futhark model}
To have a usable Futhark kernel to run in \sasmodels, it is first necessary to
define it as a Futhark program and compile it using \texttt{futhark-pyopencl}.
A model \textit{m} itself is defined as a parametric Futhark module\footnote{
  A more in-depth explanation of the complete definition of a Futhark module is 
  in \ref{sec:generating-futhark}},
which contains an Iq-function, an Iqxy- and a form\_volume function.
It is the model programmers responsibility to fill in all three functions.

@@

\subsection{Generating Futhark models using parametric modules and a header}
\label{sec:generating-futhark}
\begin{lstlisting}
module type has_iq = {
Iq: real -> real[] -> real // the model's scattering function.
Iqxy: real -> real -> real[] -> real // the model's 2D scattering function
form_volume: real[] -> real // the model's form volume function.
}  
\end{lstlisting}

% definition af gammel sasmodels-model
% evt. kald af model.

% til vægs

\subsubsection{Why not Futhark models as model info strings instead standalone modules?}
% designovervejelser, hvorfor selvstændige moduler, hvorfor ikke bare udvide kernelpy
%% hvorfor ikke bare definere futhark-Iq-funktionen som string i model.py?
% fordi der ikke er noget kodegnererering ved runtime/ingen futhark-compile-fejl
% under runtime, og prægenererede moduler kan deles med andre, uden disse
% ``andre''' behøver have futhark-oversætteren selv.

\section{Ease of use for almene brugere}
% forskel på at skrive moduler, og bare at hente et oversat modul ned

>>>>

\newcommand{\sasmodels}{\textit{SasModels}}
\newcommand{\sasview}{\textit{SasView}}
\newcommand{\iq}{\textit{I(q)}}
\newcommand{\foriq}{\texttt{for\_iq}}
\newcommand{\hasiq}{\texttt{has\_iq}}
\begin{document}

@@

\subsection{Defining a Futhark model}
\label{sec:defining-futhark-model}
To have a usable Futhark kernel to run in \sasmodels, it is first necessary to
define it as a Futhark program and compile it using \texttt{futhark-pyopencl}.
A model \textit{m} itself is defined as a parametric Futhark module\footnote{
  A more in-depth explanation of the complete definition of a Futhark module is 
  in \ref{sec:generating-futhark}},
which contains an Iq-function, an Iqxy- and a form\_volume function.
It is the model programmers responsibility to fill in all three functions.

@@

\subsection{Generating Futhark models using parametric modules and a header}
\label{sec:generating-futhark}
As mentioned in \ref{sec:defining-futhark-model}, the futhark model developer
does only need to define three functions to be able to compile a working 
\sasmodels Futhark model.
However, those three functions are merely used by a parametric Futhark module 
\texttt{for\_iq}, which takes care of the actual execution of the 
Futhark kernel.
Let's go through the actual implementation of the Futhark models.

\subsubsection{line\_futhark.fut (a Futhark model)}
\label{sec:line-fut}
\begin{figure}
  % hele line_futhark.fut
  \caption{The complete definition of the line model implemented in Futhark.}
  \label{fig:line-fut}
\end{figure}
There a several things happening in this fig. \ref{fig:line-fut}, besides the 
definition of the line Iq function.

\texttt{module line} takes either single- or double precision floats as a
parameter, and then assigns the type alias \texttt{dtype} to the type of the
input parameter. This means that we can use \texttt{dtype} in the type
declarations throughout the module, and define the module for floats in general,
instead of maybe having to define linef64- and linef32 modules.

Then, two kernels (one for each of f32 and f64) are initialised by giving
the parametric module \texttt{for\_iq} the line module and a data type 
as parameters, returning \texttt{kernel\_float32} and \texttt{kernel\_float64}.

Given these two modules, we can finally define four entry points into the
finished model. These are the four entry points that can be called by \sasmodels
whenever a \sasmodels experiment needs to execute calculations in a futhark
kernel.
Each of the entry points are simply passing on their call parameters to the 
\texttt{run\_kernel} functions defined in the \texttt{for\_iq} module.
Everything after the module line definition is boilerplate code, and could
possibly defined in a less inelegant way in a later revision, but currently, it 
is a necessary part of every Futhark model.

Other models, such as core-shell %FIX\
imports libraries so they are available in the Iq function.
%% som vist i core shell, kan det også være nødvendigt at definere og importere 
% ekstra biblioteker

\subsubsection{header/for\_iq.fut (the Futhark kernel template)}
\texttt{for\_iq.fut} contains the module type declaration for the \hasiq module,
 and it contains the definition of the parametric \foriq module.

The type declaration \hasiq ensures, that any model module that is
used as a parameter for the \foriq module defines at least the functions
Iq, Iqxy and form\_volume, as these functions are necessary for the definition
of the Futhark kernel.

Next, we have the actual definition of the module \foriq:
\foriq takes two parameters (a parametric Futhark model), and a real type 
to instantiate it with.
It then instantiates the model with the real type, making the finished Iq module
\texttt{Iq} available throughout the rest of the kernel.

\begin{figure}
  % hele for_iq.fut
  % forklar module type
  % forklar module 
  \caption{The Futhark template that is used by the individual models.}
  \label{fig:for-iq}
\end{figure}

\subsubsection{Why not implement Futhark models as model info strings 
instead of standalone modules?}
OpenCL kernels are generated by copy/pasting the desired \iq function body
into a template, and then compiling the resulting source code.
This happens at runtime in \sasmodels, which means that the kernels aren't
 actually compiled until right before \sasmodels needs them.

Futhark models eschews this design and opts instead to separate code generation
and model usage completely.
It would be simple to copy the design of the OpenCL kernels, and compile Futhark
kernels on \sasmodels runtime, but this design was cast aside in favour of the
design described in sec \ref{sec:generating-futhark}.

There are three reasons for that:
\begin{enumerate}
  \item Removing code generation from runtime unequivocally decreases the time
  it takes to run a \sasmodels calculation. Although time taken by code 
  generation is not covered by \sasmodels' built-in benchmark modules, it should
  be obvious that a program that merely executes a kernel computation has a 
  shorter runtime than a program that has to compile the kernel before it
  executes the kernel computation.

  \item Compiling a kernel just in time brings the risk of encountering 
  Futhark compilation errors at \sasmodels runtime. 
  It is arguably a user experience nightmare to encounter compilation errors at 
  runtime, which is why Futhark models keeps compilation separate from runtime.

  \item Compiling Futhark models down to self-contained\footnote{depending only 
  on numpy and pyopencl} Python libraries means that Futhark models are 
  completely portable, and as such can be used by any \sasmodels or \sasview 
  user, without them needing to have a Futhark compiler installed on their 
  workstations.
\end{enumerate}
